# Reproducibility and Validation Pipeline
---
title: "VisiumHD Reproducibility & Validation Pipeline"
author: "Michael Tuft"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      error = TRUE)
```

```{r config-input}
INPUT_SPACERANGER <- NULL
INPUT_RDS  <- "/scratch/prj/mmg_grp_single_cell_lab/projects/michael/shiny_app/data/debug_processed.rds"
BIN_SIZE   <- "square_008um"   # used only when INPUT_SPACERANGER is set
```

### Concordance validation

Set `VALIDATION_RDS` to the `clustered.rds` from the validated app session to produce a side-by-side UMAP and confusion matrix confirming 100% concordance. Set to `NULL` to skip.

```{r config-concordance}
VALIDATION_RDS <- "/scratch/prj/mmg_grp_single_cell_lab/projects/michael/shiny_app/results/063585eb11b638bc0787d8c546435d0c/clustered.rds"
OUTPUT_DIR     <- "/scratch/prj/mmg_grp_single_cell_lab/projects/michael/pipeline_output"
```

### Azimuth validation

Set `AZIMUTH_APP_RDS` to the `azimuth.rds` from the validated app session. The pipeline will run Azimuth independently and compare `predicted.class` and `predicted.subclass` labels spot-by-spot. The reference must match the one used in the app session.

```{r config-azimuth}
AZIMUTH_APP_RDS   <- "/scratch/prj/mmg_grp_single_cell_lab/projects/michael/shiny_app/results/063585eb11b638bc0787d8c546435d0c/azimuth.rds"
AZIMUTH_REFERENCE <- "mousecortexref"
```

### RCTD validation

Set `RCTD_APP_RDS` to the `rctd.rds` (or `azimuth_rctd.rds`) from the validated app session. The pipeline will run RCTD independently and compare `first_type` assignments spot-by-spot.

```{r config-rctd}
RCTD_APP_RDS       <- "/scratch/prj/mmg_grp_single_cell_lab/projects/michael/shiny_app/results/063585eb11b638bc0787d8c546435d0c/azimuth_rctd.rds"
RCTD_REFERENCE     <- "/scratch/prj/mmg_grp_single_cell_lab/projects/michael/data/zeisel_brain_ref.Rds"
RCTD_CELL_TYPE_COL <- "cell_type"
RCTD_MODE          <- "doublet"
RCTD_MAX_CORES     <- 4L
```

### BANKSY rotation validation

Set `RUN_BANKSY_ROTATION` to `TRUE` to enable Section 12. The pipeline will re-run BANKSY clustering on a version of the data with 90° rotated spatial coordinates and compare cluster assignments barcode-by-barcode against the original run. Set to `FALSE` to skip.

```{r config-banksy-rotation}
RUN_BANKSY_ROTATION <- TRUE
BANKSY_ROTATION_DEG <- 90L   # rotation angle: 90, 180, or 270

# BANKSY parameters — must match the values used in the app session
BANKSY_LAMBDA     <- 0.2    # spatial weighting (0 = transcriptomic only,
                             # 1 = fully spatial); 0.2 is the Shiny app default
BANKSY_RESOLUTION <- 0.6    # clustering resolution for BANKSY run
BANKSY_NPCS       <- 20L    # PCs used for BANKSY neighbourhood computation
```

### QC filtering thresholds

Values taken directly from the validated Shiny app session (stored in `clustered.rds@misc[["qc_params"]]`). These are the 5th and 95th percentile cutoffs of the nCounts and nFeatures distributions.

```{r config-qc}
MIN_COUNTS   <- 75     # mc_min — 5th percentile of nCount_Spatial
MAX_COUNTS   <- 580    # mc_max — 95th percentile of nCount_Spatial
MIN_FEATURES <- 61     # mf_min — 5th percentile of nFeature_Spatial
MAX_FEATURES <- 440    # mf_max — 95th percentile of nFeature_Spatial
MITO_ENABLED <- FALSE  # mt_used = NA in validated session
```

### Normalisation, PCA, and clustering

"normalize" = NormalizeData + FindVariableFeatures + ScaleData

Validated session used log-normalisation (norm_method = "normalize")

N_PCS_COMPUTE: total PCs computed by RunPCA (upper bound for elbow plot)

N_PCS_USE: PCs retained for UMAP and clustering

Validated session: n_pcs_use = 17

Validated session: resolution = 1.0, producing 18 clusters

Sketch size = 7,195 spots (10% of 71,951 post-QC spots)

```{r config-analysis}
NORM_METHOD   <- "normalize"  # NormalizeData + FindVariableFeatures + ScaleData
N_PCS_COMPUTE <- 50L          # upper bound for elbow plot inspection
N_PCS_USE     <- 17L          # validated value — set after inspecting elbow plot
RESOLUTION    <- 1.0          # validated value — produces 18 clusters
```

## Section 1 — Setup

**R package loading.** Five core packages are loaded: `Seurat` (the main single-cell/spatial analysis framework), `SeuratObject` (the underlying data structure definitions), `ggplot2` (all plotting), `patchwork` (combining multiple ggplots into one figure), and `dplyr` (data manipulation with `%>%`, `group_by`, `summarise` etc.).

**BPCells check.** `requireNamespace("BPCells", quietly = TRUE)` checks whether the BPCells package is installed and stores the result in `BPCELLS_AVAILABLE`. BPCells is the package that converts count matrices to bit-packed on-disk format so they don't need to fit entirely in RAM. It is not loaded here — just checked.

**`detect_spatial_assay`** helper function**.** This is a custom function defined here and used in sections 4 and 11. It takes a Seurat object and returns the name of the correct assay to use. The logic is: if any assay name starts with "Spatial" (case-insensitive), return that first. If not, fall back to "RNA". If neither exists, return whatever the first assay is. The function uses `Assays(obj)` to list all assays and `grep("^Spatial", ...)` to search for spatial ones. The `^` character means "starts with" in regex.

**Directory creation.** `dir.create()` creates the output directory and its `plots/` and `rds/` subdirectories. `showWarnings = FALSE` means it won't complain if they already exist. `recursive = TRUE` means it creates any missing parent directories.

### Worker constants

These must match `worker.R`

```{r setup-constants}
PIPELINE_SEED  <- 42L
SKETCH_N_CELLS <- 50000L   # target sketch size; capped at 10% of spots if smaller
```

### Output directories

```{r setup-dirs}
dir.create(OUTPUT_DIR,                     showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(OUTPUT_DIR, "plots"), showWarnings = FALSE)
dir.create(file.path(OUTPUT_DIR, "rds"),   showWarnings = FALSE)
```

### Required packages

```{r setup-packages}
required_pkgs <- c("Seurat", "SeuratObject", "ggplot2", "patchwork", "dplyr")
missing_pkgs  <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0)
  stop("Missing packages: ", paste(missing_pkgs, collapse = ", "))

library(Seurat)
library(SeuratObject)
library(ggplot2)
library(patchwork)
library(dplyr)

cat("Seurat version:", as.character(packageVersion("Seurat")), "\n")
```

### Optional packages (BPCells)

```{r setup-optional}
BPCELLS_AVAILABLE <- requireNamespace("BPCells", quietly = TRUE)
cat("BPCells available:", BPCELLS_AVAILABLE, "\n")
```

### Helper: detect spatial assay

Picks the most appropriate assay — prefers any assay starting with "Spatial", falls back to "RNA", then the first available assay. Mirrors `worker.R` lines 95–101. Picks the most appropriate assay from a Seurat object by preferring any assay whose name starts with “Spatial,” otherwise falling back to “RNA,” and if neither exists, returning the first assay in the object.

```{r setup-helper}
detect_spatial_assay <- function(obj) {
  assays  <- Assays(obj)
  spatial <- grep("^Spatial", assays, ignore.case = TRUE, value = TRUE)
  if (length(spatial) > 0) return(spatial[1])
  if ("RNA" %in% assays)   return("RNA")
  assays[1]
}
```

## Section 2 — Data Loading

```{r load-data}
# If a Spaceranger directory is provided, load raw Visium data
if (!is.null(INPUT_SPACERANGER) && nzchar(INPUT_SPACERANGER)) {

  # Construct path to the correct bin size folder (e.g., spaceranger output / spatial / 1um)
  sq_dir <- file.path(INPUT_SPACERANGER, BIN_SIZE)

  # Validate that the directory exists
  if (!dir.exists(sq_dir))
    stop("Spaceranger bin directory not found: ", sq_dir,
         "\nCheck INPUT_SPACERANGER and BIN_SIZE in CONFIG.")

  cat("Loading Spaceranger output from:", sq_dir, "\n")

  # Load 10X Visium data (counts + image + coordinates)
  obj <- Seurat::Load10X_Spatial(
    data.dir = sq_dir,
    filename = "filtered_feature_bc_matrix.h5",
    assay    = "Spatial"
  )

  cat("Raw Spaceranger data loaded.\n")

# Otherwise, if an RDS file is provided, load the pre‑processed Seurat object
} else if (!is.null(INPUT_RDS) && nzchar(INPUT_RDS)) {

  # Ensure the RDS file exists
  if (!file.exists(INPUT_RDS))
    stop("Input RDS not found: ", INPUT_RDS)

  cat("Loading pre-processed RDS:", INPUT_RDS, "\n")

  # Load the saved Seurat object
  obj <- readRDS(INPUT_RDS)

# If neither input source is provided, stop with an error
} else {
  stop("Set either INPUT_SPACERANGER or INPUT_RDS in the CONFIG section.")
}

# Summary of loaded dataset

cat("Spots loaded:   ", ncol(obj), "\n")   # number of barcodes / spots
cat("Genes detected: ", nrow(obj), "\n")   # number of genes/features
cat("Default assay:  ", DefaultAssay(obj), "\n")
cat("Assays present: ", paste(Assays(obj), collapse = ", "), "\n")
```

### Reset reductions and detect QC columns

Strips existing reductions and graphs so no stale embeddings contaminate the new run, then identifies the count and feature metadata columns. Mirrors `worker.R` `processPCA` lines 130–133. Resets reductions and graphs, simplifies the Seurat object, merges assay layers, identifies QC columns, and computes mitochondrial percentages if they are missing. From worker.R processPCA, lines 130-133

```{r load-reset}
# Reset dimensionality reductions (PCA, UMAP, etc.)
obj@reductions <- list()

# Reset neighbour graphs (SNN, KNN, etc.)
obj@graphs <- list()

# Strip object down to essential assay data and metadata
# Removes unnecessary layers, reductions, graphs, and misc slots
obj <- DietSeurat(obj, assays = DefaultAssay(obj), graphs = NULL)

# Join Seurat v5 assay layers into a single matrix (required for downstream steps)
obj <- JoinLayers(obj, assay = DefaultAssay(obj))

# Extract metadata for QC feature detection
meta <- obj@meta.data

# Identify the feature and count columns (e.g., nFeature_Spatial, nCount_Spatial)
feat_col  <- grep("^nFeature", colnames(meta), value = TRUE)[1]
count_col <- grep("^nCount",   colnames(meta), value = TRUE)[1]

cat("Count column:   ", count_col, "\n")
cat("Feature column: ", feat_col,  "\n")

# If mitochondrial percentage is required but missing, compute it now
if (!("percent.mt" %in% colnames(meta)) && MITO_ENABLED) {
  cat("percent.mt not found — computing now.\n")

  # Calculate mitochondrial percentage using gene names starting with mt- or MT-
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^mt-|^MT-")

  # Refresh metadata to include the new column
  meta <- obj@meta.data
}
```

## Section 3 — QC Filtering

### Pre-filter QC violin plot

```{r qc-prefilter-plot}
pre_features <- c(count_col, feat_col)

# If mitochondrial QC is enabled and percent.mt exists in metadata,
# include it as an additional QC feature for pre‑filter violin plots
if (MITO_ENABLED && "percent.mt" %in% colnames(meta))
  pre_features <- c(pre_features, "percent.mt")

# Number of QC panels to display
n_pre_panels <- length(pre_features)

# Build pre‑filter QC violin plots
p_pre <- VlnPlot(
    obj,
    features = pre_features,   # QC metrics to visualise
    ncol     = n_pre_panels,   # arrange all violins in one row
    pt.size  = 0               # hide individual points for cleaner violins
  ) &
  theme(
    axis.title.x = element_blank(),          # remove x‑axis titles
    axis.text.x  = element_blank(),          # hide x‑axis labels
    axis.title.y = element_text(size = 9, colour = "#444")   # style y‑axis titles
  )

# Save pre‑filter QC plot to PDF (wider if 3 panels)
ggsave(
  file.path(OUTPUT_DIR, "plots", "01_qc_prefilter.pdf"),
  plot   = p_pre,
  width  = if (n_pre_panels == 3) 10 else 7,
  height = 4
)

cat("Saved: 01_qc_prefilter.pdf\n")

# Display inline in knitted report
p_pre
```

### Apply filters

Filters cells based on count, feature, and mitochondrial thresholds, reports how many were removed, ensures enough remain, and then creates a cleaned Seurat object containing only the retained cells.

```{r qc-filter}
keep <- meta[[count_col]] >= MIN_COUNTS &
        meta[[count_col]] <= MAX_COUNTS &
        meta[[feat_col]]  >= MIN_FEATURES &
        meta[[feat_col]]  <= MAX_FEATURES

if (MITO_ENABLED && !is.na(MAX_MT) && "percent.mt" %in% colnames(meta))
  keep <- keep & (is.na(meta$percent.mt) | meta$percent.mt < MAX_MT)

keep_cells <- rownames(meta)[keep]
n_before   <- ncol(obj)
n_removed  <- n_before - length(keep_cells)

cat(sprintf("Spots before filtering: %s\n", formatC(n_before,           big.mark = ",")))
cat(sprintf("Spots after filtering:  %s\n", formatC(length(keep_cells), big.mark = ",")))
cat(sprintf("Spots removed:          %s (%.1f%%)\n",
            formatC(n_removed, big.mark = ","),
            100 * n_removed / n_before))

if (length(keep_cells) < 50)
  stop("Too few spots remain after filtering (minimum 50).")

filt <- subset(obj, cells = keep_cells)
rm(obj); gc()
```

### Post-filter QC violin plot

Selects the appropriate post‑filter QC metrics, generates violin plots for them, and saves the resulting post‑filter QC panel as a PDF.

```{r qc-postfilter-plot}
post_features <- c(count_col, feat_col)

# If mitochondrial QC is enabled and percent.mt exists, include it in the post‑filter QC plots
if (MITO_ENABLED && "percent.mt" %in% colnames(filt@meta.data))
  post_features <- c(post_features, "percent.mt")

# Number of violin plot panels to display
n_post_panels <- length(post_features)

# Build post‑filter QC violin plots

p_post <- VlnPlot(
    filt,
    features = post_features,   # QC metrics to plot
    ncol     = n_post_panels,   # arrange in one row
    pt.size  = 0                # hide individual points for cleaner violins
  ) &
  theme(
    axis.title.x = element_blank(),          # remove x‑axis titles
    axis.text.x  = element_blank(),          # hide x‑axis labels
    axis.title.y = element_text(size = 9, colour = "#444")   # style y‑axis titles
  )

# Save QC plot to PDF (wider layout if 3 panels)
ggsave(
  file.path(OUTPUT_DIR, "plots", "02_qc_postfilter.pdf"),
  plot   = p_post,
  width  = if (n_post_panels == 3) 10 else 7,
  height = 4
)

cat("Saved: 02_qc_postfilter.pdf\n")

# Display inline in knitted report
p_post

```

### BPCells on-disk conversion

```{r qc-bpcells}
if (BPCELLS_AVAILABLE) {
  cat(sprintf("Converting %s spots to BPCells on-disk storage...\n",
              formatC(ncol(filt), big.mark = ",")))
  bpcells_dir <- file.path(OUTPUT_DIR, "rds", "bpcells_counts")
  if (dir.exists(bpcells_dir)) unlink(bpcells_dir, recursive = TRUE)
  mat_disk <- BPCells::write_matrix_dir(
    mat = BPCells::convert_matrix_type(
      GetAssayData(filt, layer = "counts"), type = "uint32_t"),
    dir = bpcells_dir
  )
  filt[["RNA"]] <- SeuratObject::CreateAssayObject(counts = mat_disk)
}
```

## Section 4 — Normalisation

```{r normalisation}
assay       <- detect_spatial_assay(filt)
norm_method <- "normalize"
cat("Assay:       ", assay, "\n")
cat("Norm method: log-normalisation (NormalizeData)\n")

filt <- NormalizeData(filt, verbose = FALSE)
filt <- FindVariableFeatures(filt, selection.method = "vst",
                              nfeatures = 3000, verbose = FALSE)
filt <- ScaleData(filt, features = VariableFeatures(filt), verbose = FALSE)
gc()

cat(sprintf("Variable features: %d\n", length(VariableFeatures(filt))))
```

## Section 5 — PCA

```{r pca}
n_pcs_compute <- min(N_PCS_COMPUTE,
                     ncol(filt) - 1L,
                     length(VariableFeatures(filt)) - 1L)
if (n_pcs_compute < 5)
  stop("Not enough data to compute PCA (need at least 5 PCs).")

cat(sprintf("Computing %d PCs...\n", n_pcs_compute))

set.seed(PIPELINE_SEED)
filt <- RunPCA(filt, npcs = n_pcs_compute,
               features = VariableFeatures(filt), verbose = FALSE)
```

### Elbow plot

```{r pca-elbow}
# Compute % variance explained by each principal component:
# stdev^2 gives variance; divide by total variance and convert to %
pct_var <- filt@reductions$pca@stdev^2 /
           sum(filt@reductions$pca@stdev^2) * 100

# Number of PCs to use (set in CONFIG)
n_pcs_use <- as.integer(N_PCS_USE)

cat(sprintf("Using: %d PCs (set N_PCS_USE in CONFIG to change)\n", n_pcs_use))

# Build dataframe for plotting the elbow curve
elbow_df <- data.frame(
  pc       = seq_along(pct_var),   # PC index
  pct_var  = pct_var               # % variance explained
)

# Construct PCA elbow plot
p_elbow <- ggplot(elbow_df, aes(x = pc, y = pct_var)) +
  geom_line(colour = "#555", linewidth = 0.5) +   # line showing variance trend
  geom_point(aes(colour = pc <= n_pcs_use),       # highlight selected PCs
             size = 1.5) +
  scale_colour_manual(
    values = c("TRUE" = "#2c5f7a", "FALSE" = "#aac4d4"),
    guide  = "none"
  ) +
  geom_vline(
    xintercept = n_pcs_use,                       # vertical line at chosen PC cutoff
    linetype   = "dashed",
    colour     = "#2c5f7a",
    linewidth  = 0.7
  ) +
  annotate(
    "text",
    x     = n_pcs_use + 0.5,
    y     = max(pct_var) * 0.9,
    label = paste0("Selected: ", n_pcs_use, " PCs"),
    colour = "#2c5f7a",
    hjust  = 0,
    size   = 3.2
  ) +
  labs(
    title    = "PCA Elbow Plot",
    subtitle = "Inspect visually and set N_PCS_USE in CONFIG",
    x        = "Principal Component",
    y        = "% Variance Explained"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(colour = "#2c5f7a", face = "bold")
  )

# Save elbow plot to PDF
ggsave(
  file.path(OUTPUT_DIR, "plots", "03_elbow_plot.pdf"),
  plot   = p_elbow,
  width  = 7,
  height = 4
)

cat("Saved: 03_elbow_plot.pdf\n")

# Display inline in knitted report
p_elbow
```

### Store QC Parameters and save PCA object

```{r pca-save}
filt@misc[["qc_params"]] <- list(
  assay        = assay,
  mc_min       = MIN_COUNTS,
  mc_max       = MAX_COUNTS,
  mf_min       = MIN_FEATURES,
  mf_max       = MAX_FEATURES,
  mt_used      = if (MITO_ENABLED) MAX_MT else NA,
  n_before     = as.integer(n_before),
  n_after      = ncol(filt),
  use_sketch   = TRUE,
  norm_method  = norm_method
)

pca_rds <- file.path(OUTPUT_DIR, "rds", "pca.rds")
cat("Saving pca.rds...\n")
saveRDS(filt, pca_rds)
cat("Saved:", pca_rds)
```

## Section 6 — UMAP & Clustering

### Setup

```{r umap-setup}
params     <- filt@misc[["qc_params"]]
n_spots    <- ncol(filt)
resolution <- RESOLUTION

cat(sprintf("Spots:     %s\n", formatC(n_spots, big.mark = ",")))
cat(sprintf("PCs used:  %d\n", n_pcs_use))
cat(sprintf("Resolution: %.2f\n", resolution))
```

### Sketch-based clustering

Sketch-based pathway (worker.R processUMAP). This block sketches a subset of cells for fast clustering and UMAP, computes clusters on the sketch, projects those results back onto all cells, and standardizes the final cluster labels.

```{r umap-sketch}

# Determine how many cells to sketch:
# Use the smaller of (SKETCH_N_CELLS) or (10% of total spots)
n_sketch <- min(SKETCH_N_CELLS, round(n_spots * 0.1))

cat(sprintf("Sketching %s representative spots from %s...\n",
            formatC(n_sketch, big.mark = ","),
            formatC(n_spots,  big.mark = ",")))

# Perform Seurat sketching (LeverageScore sampling) for fast UMAP computation
set.seed(PIPELINE_SEED)
filt <- SketchData(
  object         = filt,
  ncells         = n_sketch,          # number of cells to retain in sketch
  method         = "LeverageScore",   # sampling method
  sketched.assay = "sketch"           # store sketch in a new assay
)

# Switch default assay to the sketch so downstream steps use the sketched data
DefaultAssay(filt) <- "sketch"

cat("FindNeighbors + FindClusters on sketch...\n")

# Build neighbour graph on the sketched PCA embedding
set.seed(PIPELINE_SEED)
filt <- FindNeighbors(
  filt,
  dims       = seq_len(n_pcs_use),    # PCA dimensions to use
  reduction  = "pca",
  verbose    = FALSE,
  graph.name = c("sketch_nn", "sketch_snn")   # store graphs under sketch-specific names
)

# Cluster the sketched data using the sketch SNN graph
filt <- FindClusters(
  filt,
  resolution = resolution,
  verbose    = FALSE,
  graph.name = "sketch_snn"
)

cat("RunUMAP on sketch (return.model = TRUE for projection)...\n")

# Compute UMAP on the sketch and store the model so full data can be projected later
set.seed(PIPELINE_SEED)
filt <- RunUMAP(
  filt,
  dims         = seq_len(n_pcs_use),
  return.model = TRUE,     # required for projecting full dataset onto sketch UMAP
  verbose      = FALSE
)
```

### Project to full dataset

```{r umap-project}
assay_name <- params$assay %||% "RNA"
cat(sprintf("ProjectData to all %s spots...\n",
            formatC(n_spots, big.mark = ",")))

filt <- ProjectData(
  object             = filt,
  assay              = assay_name,
  full.reduction     = "pca.full",
  sketched.assay     = "sketch",
  sketched.reduction = "pca",
  umap.model         = "umap",
  dims               = seq_len(n_pcs_use),
  refdata            = list(cluster_full = "seurat_clusters")
)
DefaultAssay(filt) <- assay_name
```

### Standardise cluster column

```{r umap-clusters}
if ("cluster_full" %in% colnames(filt@meta.data)) {
  filt@meta.data$seurat_clusters <- filt@meta.data$cluster_full
  Idents(filt) <- "seurat_clusters"
} else if ("predicted.cluster_full" %in% colnames(filt@meta.data)) {
  filt@meta.data$seurat_clusters <- filt@meta.data$predicted.cluster_full
  Idents(filt) <- "seurat_clusters"
} else {
  warning("ProjectData did not write cluster_full — non-sketch spots may have NA seurat_clusters")
}

# Update stored QC parameters
existing_params                <- filt@misc[["qc_params"]] %||% list()
existing_params$use_sketch     <- TRUE
existing_params$n_sketch       <- n_sketch
existing_params$n_pcs_use      <- n_pcs_use
existing_params$resolution     <- resolution
existing_params$cluster_method <- "sketch"
filt@misc[["qc_params"]]       <- existing_params

n_clusters <- length(unique(na.omit(as.character(filt$seurat_clusters))))
cat(sprintf("Clusters found: %d\n", n_clusters))
```

### Update QC metadata

## Section 7 — Output Plots

### UMAP coloured by cluster

```{r plots-umap}

# Extract cluster levels in numeric order (ensures consistent colouring)
lvls <- as.character(sort(as.numeric(
           unique(na.omit(as.character(filt$seurat_clusters))))))
n_cl <- length(lvls)

# Generate a base colour palette with at least 3 colours
base_cols  <- scales::hue_pal()(max(n_cl, 3))

# Map each cluster level to a unique colour
clust_cols <- setNames(base_cols[seq_len(n_cl)], lvls)

# Set Seurat cluster identities to ensure plotting uses the correct factor levels
Idents(filt) <- "seurat_clusters"
filt$seurat_clusters <- factor(as.character(filt$seurat_clusters), levels = lvls)
Idents(filt) <- "seurat_clusters"   # Reapply identities after relabelling

# Build UMAP cluster plot
p_umap <- DimPlot(
    filt,
    reduction  = "umap",     # use UMAP embedding
    label      = TRUE,       # show cluster labels on plot
    label.size = 3,
    pt.size    = 0.3,
    cols       = clust_cols  # apply consistent cluster colours
  ) +
  labs(
    title    = "UMAP — Seurat Clusters",
    subtitle = sprintf(
      "Resolution %.2f  |  %d clusters  |  %s spots",
      resolution, n_clusters, formatC(n_spots, big.mark = ",")
    )
  ) +
  theme(
    plot.title    = element_text(face = "bold", colour = "#2c5f7a"),
    plot.subtitle = element_text(colour = "#666", size = 9)
  )

# Save UMAP plot to PDF
ggsave(
  file.path(OUTPUT_DIR, "plots", "04_umap_clusters.pdf"),
  plot   = p_umap,
  width  = 7,
  height = 6
)

cat("Saved: 04_umap_clusters.pdf\n")

# Display inline in knitted report
p_umap
```

### Spatial plot coloured by cluster

```{r plots-spatial, error=TRUE, fig.height=10}
p_spatial <- tryCatch({

  # Attempt to extract Visium tissue coordinates; return NULL if unavailable
  coords <- tryCatch(GetTissueCoordinates(filt), error = function(e) NULL)

  # Stop early if no spatial coordinates exist (cannot build spatial plot)
  if (is.null(coords) || nrow(coords) == 0) stop("No tissue coordinates found.")

  # Retrieve high‑resolution image scaling factor (fallback = 1 if missing)
  sf <- tryCatch(filt@images[[1]]@scale.factors$hires / 10, error = function(e) 1)

  # Build dataframe containing scaled x/y coordinates and cluster labels
  sp_df <- data.frame(
    cell    = rownames(coords),
    x       =  coords[, 1] * sf,          # scale x‑coordinates
    y       = -coords[, 2] * sf,          # invert y to match image orientation
    cluster = factor(as.character(filt@meta.data[rownames(coords), "seurat_clusters"]),
                     levels = lvls),      # ensure consistent cluster ordering
    stringsAsFactors = FALSE
  )

  # Remove any spots lacking a cluster assignment
  sp_df <- sp_df[!is.na(sp_df$cluster), , drop = FALSE]

  # Attempt to extract the histology image and convert to a raster layer
  img_layer <- if (length(filt@images) > 0) {
    tryCatch({
      img_data <- filt@images[[1]]@image
      grid::rasterGrob(
        img_data,
        width  = grid::unit(1, "npc"),
        height = grid::unit(1, "npc"),
        interpolate = TRUE
      )
    }, error = function(e) NULL)   # If image extraction fails, skip background
  } else NULL

  # Build spatial cluster plot
  p <- ggplot(sp_df, aes(x = x, y = y, colour = cluster)) +

    # Add histology image as background if available
    { if (!is.null(img_layer))
        annotation_custom(
          img_layer,
          xmin = min(sp_df$x), xmax = max(sp_df$x),
          ymin = min(sp_df$y), ymax = max(sp_df$y)
        )
    } +

    # Plot each spot coloured by its Seurat cluster
    geom_point(size = 0.8, alpha = 0.9, stroke = 0) +

    # Use predefined cluster colour palette
    scale_colour_manual(
      values = clust_cols,
      name   = "Cluster",
      guide  = guide_legend(override.aes = list(size = 3, alpha = 1))
    ) +

    # Preserve spatial aspect ratio
    coord_fixed() +

    # Title and metadata summary
    labs(
      title    = "Spatial — Seurat Clusters",
      subtitle = sprintf(
        "%d clusters  |  resolution %.2f  |  %s spots",
        n_clusters, resolution, formatC(nrow(sp_df), big.mark = ",")
      )
    ) +

    # Remove axes and background for clean spatial view
    theme_void() +
    theme(
      plot.title      = element_text(face = "bold", colour = "#2c5f7a", size = 13),
      plot.subtitle   = element_text(colour = "#666", size = 9),
      legend.position = "right",
      legend.key.size = grid::unit(0.7, "cm"),
      legend.text     = element_text(size = 9)
    )

  p   # return plot object

}, error = function(e) {

  # If anything fails above, print message and return NULL
  cat("Spatial plot failed:", conditionMessage(e), "\n")
  NULL
})

# Save plot only if successfully generated
if (!is.null(p_spatial)) {
  ggsave(
    file.path(OUTPUT_DIR, "plots", "05_spatial_clusters.pdf"),
    plot   = p_spatial,
    width  = 10,
    height = 9,
    dpi    = 300
  )
  cat("Saved: 05_spatial_clusters.pdf\n")
  p_spatial
}
```

## Section 8 — Save Outputs

```{r save-outputs}
clustered_rds <- file.path(OUTPUT_DIR, "rds", "clustered.rds")
cat("Saving clustered.rds...\n")
saveRDS(filt, clustered_rds)
cat("Saved:", clustered_rds, "\n")

session_info_path <- file.path(OUTPUT_DIR, "session_info.txt")
writeLines(capture.output(sessionInfo()), session_info_path)
cat("Saved: session_info.txt\n")

params_final <- filt@misc[["qc_params"]]
cat(sprintf("  Input:           %s\n", INPUT_RDS %||% INPUT_SPACERANGER))
cat(sprintf("  Output dir:      %s\n", normalizePath(OUTPUT_DIR, mustWork = FALSE)))
cat(sprintf("  Spots (pre-QC):  %s\n", formatC(n_before,   big.mark = ",")))
cat(sprintf("  Spots (post-QC): %s\n", formatC(ncol(filt), big.mark = ",")))
cat(sprintf("  PCs used:        %d\n", n_pcs_use))
cat(sprintf("  Resolution:      %.2f\n", resolution))
cat(sprintf("  Clusters found:  %d\n", n_clusters))
```

## Section 9 — Concordance Validation

```{r concordance, error=TRUE, fig.width=14, fig.height=5}
if (!is.null(VALIDATION_RDS) && file.exists(VALIDATION_RDS)) {

  cat("Loading validation object:", VALIDATION_RDS, "\n")
  ref_obj      <- readRDS(VALIDATION_RDS)   # Load validated Shiny app clustering object
  shared_cells <- intersect(colnames(filt), colnames(ref_obj))  # Spots present in both objects
  n_shared     <- length(shared_cells)
  cat("Shared cells:", formatC(n_shared, big.mark = ","), "\n")

  # Extract cluster assignments for shared spots
  pipeline_cl  <- as.character(filt$seurat_clusters[shared_cells])
  ref_cl       <- as.character(ref_obj$seurat_clusters[shared_cells])

  # Compute overall cluster concordance (proportion of matching labels)
  concordance  <- mean(pipeline_cl == ref_cl)
  cat(sprintf("Concordance: %.2f%%\n", 100 * concordance))

  # UMAP panels
  # Select appropriate UMAP reduction from each object
  umap_name_pipe <- if ("umap"      %in% names(filt@reductions))     "umap"
                    else if ("full.umap" %in% names(filt@reductions)) "full.umap"
                    else names(filt@reductions)[1]

  umap_name_ref  <- if ("umap"      %in% names(ref_obj@reductions))     "umap"
                    else if ("full.umap" %in% names(ref_obj@reductions)) "full.umap"
                    else names(ref_obj@reductions)[1]

  # Extract 2D UMAP embeddings for shared cells
  emb_pipe <- as.data.frame(Embeddings(filt,    umap_name_pipe)[shared_cells, 1:2])
  emb_ref  <- as.data.frame(Embeddings(ref_obj, umap_name_ref )[shared_cells, 1:2])

  # Standardise column names
  colnames(emb_pipe) <- colnames(emb_ref) <- c("UMAP_1", "UMAP_2")

  # Attach cluster labels for colouring
  emb_pipe$cluster   <- pipeline_cl
  emb_ref$cluster    <- ref_cl

  # Build shared colour palette so clusters match visually
  all_lvls <- as.character(sort(as.numeric(unique(c(pipeline_cl, ref_cl)))))
  pal      <- setNames(scales::hue_pal()(length(all_lvls)), all_lvls)

  # Pipeline UMAP plot
  p_pipe <- ggplot(emb_pipe, aes(UMAP_1, UMAP_2, colour = cluster)) +
    geom_point(size = 0.2, alpha = 0.5) +
    scale_colour_manual(values = pal, name = "Cluster") +
    labs(title = "Reproducibility Pipeline",
         subtitle = sprintf("%s spots", formatC(n_shared, big.mark=","))) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank(), axis.text = element_blank(),
          axis.ticks = element_blank(),
          plot.title = element_text(face="bold", colour="#2c5f7a"))

  # Reference UMAP plot
  p_ref <- ggplot(emb_ref, aes(UMAP_1, UMAP_2, colour = cluster)) +
    geom_point(size = 0.2, alpha = 0.5) +
    scale_colour_manual(values = pal, name = "Cluster") +
    labs(title = "Shiny App (Validated Session)",
         subtitle = sprintf("%d clusters  |  %s spots",
                            length(all_lvls), formatC(n_shared, big.mark=","))) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank(), axis.text = element_blank(),
          axis.ticks = element_blank(),
          plot.title = element_text(face="bold", colour="#2c5f7a"))

  # Text panel summarising concordance + parameters
  p_text <- ggplot() +
    annotate("text", x = 0.5, y = 0.68, label = "Cluster concordance",
             size = 3.5, colour = "#27ae60", fontface = "bold") +
    annotate("text", x = 0.5, y = 0.52,
             label = sprintf("%.0f%%\n(%s / %s spots)",
                             100 * concordance,
                             formatC(sum(pipeline_cl == ref_cl), big.mark=","),
                             formatC(n_shared, big.mark=",")),
             size = 4.5, colour = "#27ae60", fontface = "bold") +
    annotate("text", x = 0.5, y = 0.22,
             label = sprintf("Seed: %d  |  PCs: %d  |  Resolution: %.1f\nNormalisation: %s  |  Method: sketch",
                             PIPELINE_SEED, n_pcs_use, resolution, norm_method),
             size = 2.8, colour = "#555") +
    theme_void() + xlim(0, 1) + ylim(0, 1)

  # Save combined UMAP panel
  umap_path <- file.path(OUTPUT_DIR, "plots", "06_concordance_umap.pdf")
  ggsave(umap_path,
         plot   = p_pipe + p_ref + p_text +
                  patchwork::plot_layout(ncol = 3, widths = c(2, 2, 1.2)),
         device = "pdf", width = 16, height = 5.5, dpi = 300)
  cat("Saved:", umap_path, "\n")

  # Confusion matrix
  # Define consistent cluster level ordering
  lvls  <- as.character(sort(as.numeric(unique(c(pipeline_cl, ref_cl)))))

  # Build contingency table of pipeline vs reference clusters
  cm    <- table(Pipeline = factor(pipeline_cl, levels = lvls),
                 App      = factor(ref_cl,      levels = lvls))
  cm_df <- as.data.frame(cm)

  # Heatmap of cluster agreement
  p_cm <- ggplot(cm_df, aes(x = App, y = Pipeline, fill = Freq)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(data = cm_df[cm_df$Freq > 0, ],
              aes(label = formatC(Freq, big.mark=",")),
              size = 2.5, colour = "white", fontface = "bold") +
    scale_fill_gradient(low = "#f0f4f8", high = "#2c5f7a",
                        name = "Spots", trans = "log1p",
                        labels = scales::comma) +
    labs(title    = sprintf("Cluster Concordance Matrix (%.0f%%)", 100 * concordance),
         subtitle = "Pipeline (rows) vs Shiny App validated session (columns)",
         x = "Shiny App cluster", y = "Pipeline cluster") +
    theme_minimal(base_size = 10) +
    theme(panel.grid    = element_blank(),
          axis.text     = element_text(size = 8),
          plot.title    = element_text(face = "bold", colour = "#2c5f7a"),
          plot.subtitle = element_text(colour = "#666", size = 8))

  # Save confusion matrix
  cm_path <- file.path(OUTPUT_DIR, "plots", "07_concordance_matrix.pdf")
  ggsave(cm_path, plot = p_cm, device = "pdf", width = 8, height = 7, dpi = 300)
  cat("Saved:", cm_path, "\n")

  # Print final concordance summary
  cat(sprintf("Concordance result: %.2f%% (%s/%s spots)\n",
              100 * concordance,
              formatC(sum(pipeline_cl == ref_cl), big.mark=","),
              formatC(n_shared, big.mark=",")))

  # Display inline in knitted report
  (p_pipe + p_ref + p_text + patchwork::plot_layout(ncol = 3, widths = c(2, 2, 1.2)))

} else {
  # Skip section if validation file missing
  cat("\nSection 9 skipped — VALIDATION_RDS not set or file not found.\n")
}
```

## Section 10 — Azimuth Reproducibility Validation

```{r azimuth-validation, error=TRUE, fig.width=12, fig.height=9}
if (!is.null(AZIMUTH_APP_RDS) && file.exists(AZIMUTH_APP_RDS) &&
    requireNamespace("Azimuth", quietly = TRUE)) {

  cat("Running Azimuth on pipeline clustered object...\n")

  # Switch to sequential execution — required to avoid future globals size limit
  old_plan <- future::plan()
  on.exit(future::plan(old_plan), add = TRUE)
  future::plan(future::sequential)
  options(future.globals.maxSize = +Inf)

  # Join layers for Assay5 objects (Seurat v5)
  az_obj <- filt
  for (az_assay in intersect(Assays(az_obj), c("RNA", "Spatial"))) {
    tryCatch({
      if (inherits(az_obj[[az_assay]], "Assay5"))
        az_obj <- JoinLayers(az_obj, assay = az_assay)
    }, error = function(e) message("[azimuth] JoinLayers skipped: ", e$message))
  }

  # Run Azimuth
  query_assay <- if ("Spatial" %in% Assays(az_obj)) "Spatial" else DefaultAssay(az_obj)
  az_obj <- Azimuth::RunAzimuth(query = az_obj, reference = AZIMUTH_REFERENCE,
                                 assay = query_assay, verbose = FALSE)
  cat("Azimuth pipeline run complete.\n")

  # Load validated app Azimuth results
  cat("Loading app Azimuth results from:", AZIMUTH_APP_RDS, "\n")
  app_az    <- readRDS(AZIMUTH_APP_RDS)
  shared_az <- intersect(colnames(az_obj), colnames(app_az))
  cat("Shared spots:", formatC(length(shared_az), big.mark=","), "\n")

  # Helper: concordance at any Azimuth label level
  # Arguments passed explicitly to avoid R scoping issues with az_obj/app_az
  run_az_concordance <- function(pipeline_obj, app_obj, shared,
                                  label_col, level_name, out_prefix, fig_num,
                                  fig_width = 10, fig_height = 8, text_size = 8) {

    if (!(label_col %in% colnames(pipeline_obj@meta.data)) ||
        !(label_col %in% colnames(app_obj@meta.data))) {
      cat(sprintf("Skipping %s concordance — column '%s' not found.\n",
                  level_name, label_col))
      return(invisible(NULL))
    }

    pipeline_l  <- as.character(pipeline_obj@meta.data[shared, label_col])
    app_l       <- as.character(app_obj@meta.data[shared, label_col])
    both_v      <- !is.na(pipeline_l) & !is.na(app_l)
    n_v         <- sum(both_v)
    concordance <- mean(pipeline_l[both_v] == app_l[both_v])

    cat(sprintf("\nAzimuth %s concordance: %.2f%% (%s/%s valid spots)\n",
                level_name, 100 * concordance,
                formatC(sum(pipeline_l[both_v] == app_l[both_v]), big.mark=","),
                formatC(n_v, big.mark=",")))

    df <- data.frame(pipeline = pipeline_l[both_v], app = app_l[both_v],
                     match    = pipeline_l[both_v] == app_l[both_v])

    per_label <- df %>%
      dplyr::group_by(app) %>%
      dplyr::summarise(n_total   = dplyr::n(),
                       n_match   = sum(match),
                       pct_match = round(100 * mean(match), 1),
                       .groups   = "drop") %>%
      dplyr::arrange(dplyr::desc(pct_match))

    cat(sprintf("Per-%s concordance:\n", level_name))
    print(as.data.frame(per_label), row.names = FALSE)

    write.csv(per_label,
              file.path(OUTPUT_DIR, "plots",
                        paste0("azimuth_", out_prefix, "_concordance.csv")),
              row.names = FALSE)
    cat(sprintf("Saved: azimuth_%s_concordance.csv\n", out_prefix))

    cross    <- table(Pipeline = df$pipeline, App = df$app)
    cross_df <- as.data.frame(cross) %>%
      dplyr::group_by(App) %>%
      dplyr::mutate(prop = Freq / sum(Freq)) %>%
      dplyr::ungroup()

    p <- ggplot(cross_df, aes(x = App, y = Pipeline, fill = prop)) +
      geom_tile(colour = "white", linewidth = 0.3) +
      geom_text(data = cross_df[cross_df$prop >= 0.05, ],
                aes(label = sprintf("%.0f%%", 100 * prop)),
                size = 2.8, colour = "white", fontface = "bold") +
      scale_fill_gradient(low = "#f0f4f8", high = "#2c5f7a",
                          name = "Proportion", labels = scales::percent) +
      labs(title    = sprintf("Azimuth %s Reproducibility — %.1f%% concordance",
                              level_name, 100 * concordance),
           subtitle = sprintf("Row proportions: proportion of each App %s reproduced by Pipeline",
                              level_name),
           x = sprintf("Shiny App %s", label_col),
           y = sprintf("Pipeline %s",  label_col)) +
      theme_minimal(base_size = 9) +
      theme(axis.text.x   = element_text(angle = 45, hjust = 1, size = text_size),
            axis.text.y   = element_text(size = text_size),
            panel.grid    = element_blank(),
            plot.title    = element_text(face = "bold", colour = "#2c5f7a"),
            plot.subtitle = element_text(colour = "#666", size = 8))

    out_path <- file.path(OUTPUT_DIR, "plots",
                          sprintf("%s_azimuth_%s_concordance.pdf",
                                  fig_num, out_prefix))
    ggsave(out_path, plot = p, device = "pdf",
           width = fig_width, height = fig_height, dpi = 300)
    cat("Saved:", out_path, "\n")

    
    cat(sprintf("Azimuth %s Concordance: %.2f%% (%s/%s spots)\n",
                toupper(level_name), 100 * concordance,
                formatC(sum(pipeline_l[both_v] == app_l[both_v]), big.mark=","),
                formatC(n_v, big.mark=",")))

    print(p)
    invisible(concordance)
  }

  # Run at class level (GABAergic / Glutamatergic / Non-Neuronal)
  run_az_concordance(az_obj, app_az, shared_az,
                     "predicted.class", "class", "class",
                     "08a", fig_width = 8, fig_height = 7)

  # Run at subclass level (L5 IT, Oligo, Micro-PVM etc.)
  run_az_concordance(az_obj, app_az, shared_az,
                     "predicted.subclass", "subclass", "subclass",
                     "08b", fig_width = 14, fig_height = 12, text_size = 7)

} else {
  cat("\nSection 10 skipped — AZIMUTH_APP_RDS not set, file not found, or Azimuth not installed.\n")
}
```

## Section 11 — RCTD Reproducibility Validation

```{r rctd-validation, error=TRUE, fig.width=9, fig.height=8, message=FALSE}
if (!is.null(RCTD_APP_RDS) && file.exists(RCTD_APP_RDS) &&
    requireNamespace("spacexr", quietly = TRUE)) {

  cat("Preparing spatial data for RCTD...\n")

  sp_assay  <- if ("Spatial" %in% Assays(filt)) "Spatial" else DefaultAssay(filt)
  sp_counts <- tryCatch(
    LayerData(filt, assay = sp_assay, layer = "counts"),
    error = function(e) GetAssayData(filt, assay = sp_assay, slot = "counts")
  )
  if (!inherits(sp_counts, "dgCMatrix"))
    sp_counts <- as(sp_counts, "CsparseMatrix")

  coords    <- GetTissueCoordinates(filt)
  coord_df  <- data.frame(x = coords[, 1], y = coords[, 2],
                           row.names = rownames(coords))
  shared_cells <- intersect(colnames(sp_counts), rownames(coord_df))
  sp_counts <- sp_counts[, shared_cells, drop = FALSE]
  coord_df  <- coord_df[shared_cells,  , drop = FALSE]
  puck      <- spacexr::SpatialRNA(coord_df, sp_counts, colSums(sp_counts))
  cat(sprintf("SpatialRNA: %d spots, %d genes\n", ncol(sp_counts), nrow(sp_counts)))

  # Load and prepare reference
  cat("Loading RCTD reference:", RCTD_REFERENCE, "\n")
  ref_obj    <- readRDS(RCTD_REFERENCE)
  ref_assay  <- if ("RNA" %in% Assays(ref_obj)) "RNA" else DefaultAssay(ref_obj)
  ref_counts <- tryCatch(
    LayerData(ref_obj, assay = ref_assay, layer = "counts"),
    error = function(e) GetAssayData(ref_obj, assay = ref_assay, slot = "counts")
  )
  if (!inherits(ref_counts, "dgCMatrix"))
    ref_counts <- as(ref_counts, "CsparseMatrix")
  
  # Takes the raw cell‑type labels from the reference object
  cell_types_raw <- ref_obj@meta.data[[RCTD_CELL_TYPE_COL]]
  names(cell_types_raw) <- colnames(ref_obj)
  # Drops any cells with missing labels
  valid_cells  <- !is.na(cell_types_raw) & nzchar(as.character(cell_types_raw))
  # Aligns the remaining labels to the filtered count matrix
  ref_counts   <- ref_counts[, valid_cells, drop = FALSE]
  # Returns a clean factor of valid cell types for downstream analysis
  cell_types   <- as.factor(cell_types_raw[valid_cells])

  # Downsample to 500 cells per type (mirrors worker.R)
  set.seed(PIPELINE_SEED)
  keep_cells <- unlist(lapply(levels(cell_types), function(ct) {
    idx <- which(cell_types == ct)
    if (length(idx) > 500L) sample(idx, 500L) else idx
  }))
  ref_counts <- ref_counts[, keep_cells, drop = FALSE]
  cell_types <- droplevels(cell_types[keep_cells])

  # Remove types with < 25 cells
  ct_counts  <- table(cell_types)
  keep_types <- names(ct_counts)[ct_counts >= 25]
  keep_ref   <- cell_types %in% keep_types
  ref_counts <- ref_counts[, keep_ref, drop = FALSE]
  cell_types <- droplevels(cell_types[keep_ref])

  reference <- spacexr::Reference(ref_counts, cell_types, colSums(ref_counts))
  cat(sprintf("Reference: %d cells, %d types\n",
              ncol(ref_counts), length(levels(cell_types))))

  # Run RCTD
  cat(sprintf("Running RCTD (%s mode, %d cores)...\n", RCTD_MODE, RCTD_MAX_CORES))
  set.seed(PIPELINE_SEED)
  RCTD_run      <- spacexr::create.RCTD(puck, reference, max_cores = RCTD_MAX_CORES)
  RCTD_run      <- spacexr::run.RCTD(RCTD_run, doublet_mode = RCTD_MODE)
  pipeline_rctd <- AddMetaData(filt, metadata = RCTD_run@results$results_df)
  cat("RCTD pipeline run complete.\n")

  # Load app RCTD results and compare
  cat("Loading app RCTD results from:", RCTD_APP_RDS, "\n")
  app_rctd     <- readRDS(RCTD_APP_RDS)
  shared_rctd  <- intersect(colnames(pipeline_rctd), colnames(app_rctd))
  cat("Shared spots:", formatC(length(shared_rctd), big.mark=","), "\n")

  pipeline_types   <- as.character(pipeline_rctd$first_type[shared_rctd])
  app_types        <- as.character(app_rctd$first_type[shared_rctd])
  both_valid_r     <- !is.na(pipeline_types) & !is.na(app_types)
  n_valid_r        <- sum(both_valid_r)
  rctd_concordance <- mean(pipeline_types[both_valid_r] == app_types[both_valid_r])

  cat(sprintf("RCTD first_type concordance: %.2f%% (%s/%s valid spots)\n",
              100 * rctd_concordance,
              formatC(sum(pipeline_types[both_valid_r] == app_types[both_valid_r]),
                      big.mark=","),
              formatC(n_valid_r, big.mark=",")))

  rctd_df <- data.frame(
    pipeline = pipeline_types[both_valid_r],
    app      = app_types[both_valid_r],
    match    = pipeline_types[both_valid_r] == app_types[both_valid_r]
  )

  per_type <- rctd_df %>%
    dplyr::group_by(app) %>%
    dplyr::summarise(n_total   = dplyr::n(),
                     n_match   = sum(match),
                     pct_match = round(100 * mean(match), 1),
                     .groups   = "drop") %>%
    dplyr::arrange(dplyr::desc(pct_match))

  cat("\nPer-type RCTD concordance:\n")
  print(as.data.frame(per_type), row.names = FALSE)

  write.csv(per_type, file.path(OUTPUT_DIR, "plots", "rctd_per_type_concordance.csv"),
            row.names = FALSE)
  cat("Saved: rctd_per_type_concordance.csv\n")

  cross_rctd    <- table(Pipeline = rctd_df$pipeline, App = rctd_df$app)
  cross_df_rctd <- as.data.frame(cross_rctd) %>%
    dplyr::group_by(App) %>%
    dplyr::mutate(prop = Freq / sum(Freq)) %>%
    dplyr::ungroup()

  p_rctd <- ggplot(cross_df_rctd, aes(x = App, y = Pipeline, fill = prop)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(data = cross_df_rctd[cross_df_rctd$prop >= 0.05, ],
              aes(label = sprintf("%.0f%%", 100 * prop)),
              size = 3, colour = "white", fontface = "bold") +
    scale_fill_gradient(low = "#f0f4f8", high = "#2c5f7a",
                        name = "Proportion", labels = scales::percent) +
    labs(title    = sprintf("RCTD Reproducibility — %.1f%% concordance",
                            100 * rctd_concordance),
         subtitle = "Row proportions: proportion of each App cell type reproduced by Pipeline",
         x = "Shiny App first_type", y = "Pipeline first_type") +
    theme_minimal(base_size = 10) +
    theme(axis.text.x   = element_text(angle = 30, hjust = 1, size = 9),
          axis.text.y   = element_text(size = 9),
          panel.grid    = element_blank(),
          plot.title    = element_text(face = "bold", colour = "#2c5f7a"),
          plot.subtitle = element_text(colour = "#666", size = 8))

  rctd_path <- file.path(OUTPUT_DIR, "plots", "09_rctd_concordance.pdf")
  ggsave(rctd_path, plot = p_rctd, device = "pdf", width = 8, height = 7, dpi = 300)
  cat("Saved:", rctd_path, "\n")

  cat(sprintf("RCTD Concordance: %.2f%% (%s/%s valid spots)\n",
              100 * rctd_concordance,
              formatC(sum(pipeline_types[both_valid_r] == app_types[both_valid_r]),
                      big.mark=","),
              formatC(n_valid_r, big.mark=",")))

  p_rctd

} else {
  cat("\nSection 11 skipped — RCTD_APP_RDS not set, file not found, or spacexr not installed.\n")
}
```

## Section 12 — BANKSY Rotation Validation

```{r banksy-rotation-validation, error=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
if (isTRUE(RUN_BANKSY_ROTATION) &&
    requireNamespace("Banksy",               quietly = TRUE) &&
    requireNamespace("SingleCellExperiment", quietly = TRUE) &&
    requireNamespace("SummarizedExperiment", quietly = TRUE) &&
    requireNamespace("S4Vectors",            quietly = TRUE) &&
    requireNamespace("plotly",               quietly = TRUE) &&
    requireNamespace("mclust",               quietly = TRUE)) {

  ROTATION_DEGREES <- BANKSY_ROTATION_DEG

  cat(sprintf("Rotation: %d degrees\n", ROTATION_DEGREES))

  # Helper: rotate x/y vectors around their joint centroid
  rotate_coords_banksy <- function(x, y, degrees) {
    deg <- as.integer(degrees) %% 360L
    if (deg == 0L) return(list(x = x, y = y))
    cx <- (max(x) + min(x)) / 2
    cy <- (max(y) + min(y)) / 2
    dx <- x - cx; dy <- y - cy
    if (deg == 90L)  return(list(x = cx + dy,  y = cy - dx))
    if (deg == 180L) return(list(x = cx - dx,  y = cy - dy))
    list(x = cx - dy, y = cy + dx)  # 270
  }

  run_banksy_pipeline <- function(obj_input, label = "original",
                                  coords_override = NULL) {
    cat(sprintf("Running BANKSY (%s)\n", label))

    obj <- obj_input

    # Ensure layers are joined for Banksy
    for (a in intersect(Assays(obj), c("RNA", "Spatial"))) {
      tryCatch({
        if (inherits(obj[[a]], "Assay5"))
          obj <- JoinLayers(obj, assay = a)
      }, error = function(e) NULL)
    }

    # Use supplied coordinates if given, else extract from the object
    if (!is.null(coords_override)) {
      coords <- coords_override
    } else {
      coords <- tryCatch(GetTissueCoordinates(obj), error = function(e) NULL)
    }
    if (is.null(coords) || nrow(coords) == 0)
      stop("No tissue coordinates found on object.")

    # Standardise coordinate names
    if ("imagecol" %in% colnames(coords) && !"x" %in% colnames(coords))
      colnames(coords)[colnames(coords) == "imagecol"] <- "x"
    if ("imagerow" %in% colnames(coords) && !"y" %in% colnames(coords))
      colnames(coords)[colnames(coords) == "imagerow"] <- "y"

    # Match coordinates to object cells
    shared_bc_b <- intersect(rownames(coords), colnames(obj))
    coords      <- coords[shared_bc_b, , drop = FALSE]
    obj_sub     <- obj[, shared_bc_b]

    # Convert Seurat to SingleCellExperiment
    assay_use <- if ("Spatial" %in% Assays(obj_sub)) "Spatial" else "RNA"

    counts_mat <- tryCatch(
      LayerData(obj_sub, assay = assay_use, layer = "counts"),
      error = function(e) GetAssayData(obj_sub, assay = assay_use, slot = "counts")
    )
    if (!inherits(counts_mat, "dgCMatrix"))
      counts_mat <- as(counts_mat, "CsparseMatrix")

    norm_mat <- tryCatch(
      LayerData(obj_sub, assay = assay_use, layer = "data"),
      error = function(e) GetAssayData(obj_sub, assay = assay_use, slot = "data")
    )
    if (!inherits(norm_mat, "dgCMatrix"))
      norm_mat <- as(norm_mat, "CsparseMatrix")
    # computeBanksy expects coordinate columns named "sdimx" and "sdimy" by default
    sce <- SingleCellExperiment::SingleCellExperiment(
      assays   = list(counts = counts_mat, logcounts = norm_mat),
      colData  = S4Vectors::DataFrame(
        sdimx = coords[colnames(counts_mat), "x"],
        sdimy = coords[colnames(counts_mat), "y"]
      )
    )

    # Step 1: computeBanksy
    set.seed(PIPELINE_SEED)
    sce <- Banksy::computeBanksy(
      se           = sce,
      assay_name   = "logcounts",
      compute_agf  = FALSE,
      k_geom       = 6L,
      spatial_mode = "kNN_r",
      M            = 1L,
      coord_names  = c("sdimx", "sdimy"),
      verbose      = FALSE
    )

    # Step 2: runBanksyPCA
    set.seed(PIPELINE_SEED)
    sce <- Banksy::runBanksyPCA(
      sce,
      use_agf   = FALSE,
      lambda    = BANKSY_LAMBDA,
      npcs      = BANKSY_NPCS,
      seed      = PIPELINE_SEED
    )

    # Step 3: clusterBanksy
    set.seed(PIPELINE_SEED)
    sce <- Banksy::clusterBanksy(
      sce,
      use_agf    = FALSE,
      lambda     = BANKSY_LAMBDA,
      resolution = BANKSY_RESOLUTION,
      seed       = PIPELINE_SEED,
      verbose    = FALSE
    )

    # Extract cluster column
    clust_cols <- grep("^clust_", colnames(SummarizedExperiment::colData(sce)),
                       value = TRUE)
    if (length(clust_cols) == 0)
      stop("No Banksy cluster column found in colData.")
    clust_col <- clust_cols[length(clust_cols)]

    clusters <- setNames(
      as.character(SummarizedExperiment::colData(sce)[[clust_col]]),
      colnames(sce)
    )
    n_cl <- length(unique(clusters))
    cat(sprintf("  Clusters found: %d  (column: %s)\n", n_cl, clust_col))
    clusters
  }

  # Run 1: Original coordinates
  cl_original <- run_banksy_pipeline(filt, label = "original")

  # Run 2: Rotated coordinates
  # Rotate the coordinate table directly and pass it via coords_override,
  # avoiding any image-slot manipulation (VisiumV2 has no @coordinates slot).
  coords_orig <- GetTissueCoordinates(filt)

  # Standardise names to x / y
  if ("imagecol" %in% colnames(coords_orig) && !"x" %in% colnames(coords_orig))
    colnames(coords_orig)[colnames(coords_orig) == "imagecol"] <- "x"
  if ("imagerow" %in% colnames(coords_orig) && !"y" %in% colnames(coords_orig))
    colnames(coords_orig)[colnames(coords_orig) == "imagerow"] <- "y"

  rotated <- rotate_coords_banksy(coords_orig$x, coords_orig$y, ROTATION_DEGREES)

  coords_rot <- coords_orig
  coords_rot$x <- rotated$x
  coords_rot$y <- rotated$y
  cat(sprintf("Coordinates rotated by %d degrees.\n", ROTATION_DEGREES))

  cl_rotated <- run_banksy_pipeline(
    filt,
    label           = sprintf("%d-degree rotation", ROTATION_DEGREES),
    coords_override = coords_rot
  )

  # Step 3: Compare barcode-by-barcode
  shared_bc <- intersect(names(cl_original), names(cl_rotated))
  cat(sprintf("\nShared barcodes: %s\n", formatC(length(shared_bc), big.mark=",")))

  orig_cl <- cl_original[shared_bc]
  rot_cl  <- cl_rotated[shared_bc]

  pct_identical <- mean(orig_cl == rot_cl) * 100
  ari_val <- mclust::adjustedRandIndex(orig_cl, rot_cl)

  cat(sprintf("Identical assignments: %.2f%% (%s/%s barcodes)\n",
              pct_identical,
              formatC(sum(orig_cl == rot_cl), big.mark=","),
              formatC(length(shared_bc), big.mark=",")))
  cat(sprintf("Adjusted Rand Index:   %.4f\n", ari_val))

  # Save concordance CSV
  concordance_df <- data.frame(
    barcode          = shared_bc,
    cluster_original = orig_cl,
    cluster_rotated  = rot_cl,
    match            = orig_cl == rot_cl,
    stringsAsFactors = FALSE
  )
  conc_csv <- file.path(OUTPUT_DIR, "plots",
                        "12_banksy_rotation_concordance.csv")
  write.csv(concordance_df, conc_csv, row.names = FALSE)
  cat("Saved:", conc_csv, "\n")

  # Confusion heatmap
  cross_b    <- table(Original = orig_cl, Rotated = rot_cl)
  cross_df_b <- as.data.frame(cross_b) %>%
    dplyr::group_by(Original) %>%
    dplyr::mutate(prop = Freq / sum(Freq)) %>%
    dplyr::ungroup()

  lvls_b <- as.character(
    sort(as.numeric(unique(c(as.character(cross_df_b$Original),
                             as.character(cross_df_b$Rotated))))))

  cross_df_b$Original <- factor(cross_df_b$Original, levels = lvls_b)
  cross_df_b$Rotated  <- factor(cross_df_b$Rotated,  levels = lvls_b)

  p_banksy_heat <- ggplot(cross_df_b, aes(x = Rotated, y = Original,
                                          fill = prop)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_text(data = cross_df_b[cross_df_b$prop >= 0.05, ],
              aes(label = sprintf("%.0f%%", 100 * prop)),
              size = 2.8, colour = "white", fontface = "bold") +
    scale_fill_gradient(low = "#f0f4f8", high = "#2c5f7a",
                        name = "Proportion", labels = scales::percent) +
    labs(title    = sprintf("BANKSY Rotation Concordance — %.1f%% identical  |  ARI = %.4f",
                            pct_identical, ari_val),
         subtitle = sprintf("Original (rows) vs %d°-rotated (columns) cluster assignments",
                            ROTATION_DEGREES),
         x = sprintf("%d° Rotated cluster", ROTATION_DEGREES),
         y = "Original cluster") +
    theme_minimal(base_size = 10) +
    theme(axis.text.x   = element_text(size = 8),
          axis.text.y   = element_text(size = 8),
          panel.grid    = element_blank(),
          plot.title    = element_text(face = "bold", colour = "#2c5f7a"),
          plot.subtitle = element_text(colour = "#666", size = 8))

  heat_path <- file.path(OUTPUT_DIR, "plots",
                         "12_banksy_rotation_heatmap.pdf")
  ggsave(heat_path, plot = p_banksy_heat,
         device = "pdf", width = 9, height = 8, dpi = 300)
  cat("Saved:", heat_path, "\n")
  print(p_banksy_heat)

  # Interactive Plotly Venn Diagram
  cat("\nBuilding interactive Plotly cluster overlap diagram...\n")

  orig_levels <- as.character(sort(as.numeric(unique(orig_cl))))
  rot_levels  <- as.character(sort(as.numeric(unique(rot_cl))))

  overlap_list <- list()
  for (oc in orig_levels) {
    for (rc in rot_levels) {
      n <- sum(orig_cl == oc & rot_cl == rc)
      if (n > 0) {
        overlap_list[[length(overlap_list) + 1]] <- data.frame(
          orig_cluster = oc,
          rot_cluster  = rc,
          n_barcodes   = n,
          stringsAsFactors = FALSE
        )
      }
    }
  }
  overlap_df <- do.call(rbind, overlap_list)
  overlap_df$orig_num <- as.numeric(overlap_df$orig_cluster)
  overlap_df$rot_num  <- as.numeric(overlap_df$rot_cluster)
  overlap_df$is_diag  <- overlap_df$orig_cluster == overlap_df$rot_cluster

  overlap_df$colour <- ifelse(overlap_df$is_diag, "#2c5f7a", "#e07b5a")
  overlap_df$label  <- sprintf(
    "Original: %s\nRotated: %s\nBarcodes: %s",
    overlap_df$orig_cluster,
    overlap_df$rot_cluster,
    formatC(overlap_df$n_barcodes, big.mark = ",")
  )

  max_n <- max(overlap_df$n_barcodes)
  overlap_df$size <- 6 + 44 * sqrt(overlap_df$n_barcodes / max_n)

  # Concordance venn diagram
  fig_venn <- plotly::plot_ly(
    data   = overlap_df,
    x      = ~rot_num,
    y      = ~orig_num,
    type   = "scatter",
    mode   = "markers",
    marker = list(
      size    = ~size,
      color   = ~colour,
      opacity = 0.75,
      line    = list(color = "white", width = 1)
    ),
    text      = ~label,
    hoverinfo = "text"
  ) %>%
    plotly::layout(
      title = list(
        text = sprintf(
          "<b>BANKSY Cluster Overlap: Original vs %d° Rotated</b><br><sup>Bubble size = barcode count. Teal = matched cluster. Coral = cross-cluster overlap.</sup>",
          ROTATION_DEGREES),
        font = list(size = 14)
      ),
      xaxis = list(
        title      = sprintf("%d° Rotated cluster", ROTATION_DEGREES),
        tickmode   = "array",
        tickvals   = as.numeric(rot_levels),
        ticktext   = rot_levels,
        showgrid   = TRUE,
        gridcolor  = "#eee"
      ),
      yaxis = list(
        title      = "Original cluster",
        tickmode   = "array",
        tickvals   = as.numeric(orig_levels),
        ticktext   = orig_levels,
        showgrid   = TRUE,
        gridcolor  = "#eee",
        autorange  = "reversed"
      ),
      plot_bgcolor  = "#fafafa",
      paper_bgcolor = "#ffffff",
      hoverlabel    = list(
        bgcolor  = "white",
        font     = list(size = 12)
      )
    ) %>%
    plotly::add_annotations(
      text      = sprintf("ARI = %.4f | %.1f%% identical", ari_val, pct_identical),
      x         = 0.5, y = -0.12,
      xref      = "paper", yref = "paper",
      showarrow = FALSE,
      font      = list(size = 11, color = "#555")
    )

  print(fig_venn)

  # Summary
  cat(sprintf("BANKSY rotation validation:\n"))
  cat(sprintf("  Rotation applied:    %d degrees\n", ROTATION_DEGREES))
  cat(sprintf("  Shared barcodes:     %s\n",
              formatC(length(shared_bc), big.mark=",")))
  cat(sprintf("  Identical clusters:  %.2f%%\n", pct_identical))
  cat(sprintf("  Adjusted Rand Index: %.4f\n",  ari_val))
  cat(sprintf("  Interpretation:      %s\n",
              if (ari_val > 0.99) "BANKSY is fully orientation-invariant"
              else if (ari_val > 0.90) "Minor label permutations only"
              else "Unexpected sensitivity to coordinate orientation"))

} else {
  cat("\nSection 12 skipped — RUN_BANKSY_ROTATION is FALSE, or Banksy/SummarizedExperiment/plotly/mclust not installed.\n")
  cat("Install with:\n")
  cat("  BiocManager::install(c('Banksy', 'SummarizedExperiment'))\n")
  cat("  install.packages(c('plotly', 'mclust'))\n")
}
```
